---
title: "BeeResp_15Feb"
author: "Callin Switzer"
date: "February 15, 2017; Update 1 March 2017; Update March 16"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

### Summary:
Two avenues of analysis:

1.  First we see how $\Delta$load affects $\Delta$ArcLength^2 and $\Delta$frequency^2, accounting for bee size and treatment order
2.  We check to see how $\Delta$frequency and $\Delta$ArcLength are associated with $\Delta$MetabolicRate, while accounting for $\Delta$load

### Measured variables:
- Bee ID: Name given to an invididual (subscript in equations)
- Order: This is the trial number per individual, either 1 or 2
- Treatment: Either light or heavy- H is with both nectar and external load, L is just nectar load (subscript in equations)
- MStarved: The empty mass of the bee after being starved for days, until it no longer buzzes when prodded in grams
- M2: Total mass of the bee and load at beginning of trial in grams
- MF: The mass of the bee and load at end of trial in grams
- ITspan=intertegular span in m
- S=area of both forewings in $m^2$
- MetR= Metabolic rate in mL $CO_2$ $hr^{-1}$
- freq= wing beat frequency in Hz
- amp= stroke amplitude of forewing in degrees
- wingLen = length of forewing from wing base to tip in m
 
 
### Calculated variables:
- MT: Mtrial, total mass, =(M2+MF)/2 = the average mass of the course of the trial
- Load= Mtrial- Mstarved
- Perload=  Percent load, %load (Load/Mstarved)x100
- ArcL= (.75 $\cdotp$ wlength) (Amp ($\pi$/180)) 	pi/180 converts to radians
- U= wing velocity, arcL$\cdotp$ frq $\cdotp$ 2   (the 2 comes from the fact that frequency consists of both an up and down stroke, vel=∆x/∆t, ∆x=arclength, ∆t=1/(freg $\cdotp$ 2)
- Freqsq=$Freq^2$
- ArcLsq=$ArcL^2$
- Force=$U^2 \cdotp S$
- Mass_Spec_MetR = Mass specific metabolic rate. MetR / MT


---
## Install packages and read data
---

```{r, message=FALSE, warning=FALSE}
# install packages
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if(length(new.pkg)) install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

packages <- c("lme4", "influence.ME", "data.table", "tidyverse")
ipak(packages)

# set ggplot theme
theme_set(theme_classic() + 
            theme(axis.text=element_text(colour="black"), 
                  text=element_text(size=10)))

# set  directories
dataDir <- file.path(getwd(), "data")
figDir <- file.path(getwd(), "figures")
dataOut <- file.path(getwd(), "dataOutput")

print(paste("last run ", Sys.time()))
print(R.version)

```

---
# Import data
---

```{r}
bdta <- read_csv(file.path(dataDir, "FinalBeeRespData_raw.csv"))
summary(bdta)
```

---
# Data tidying
---

```{r}
bdta <- bdta %>%
  mutate(
     MT = (M2 + MF)/2,
     load = MT - Mstarved,
     perLoad = load / Mstarved * 100,
     arcL = (0.75 *wingLen) * (amp * (pi / 180)),
     U = arcL * freq * 2,
     freq2 = freq^2,
     arcL2 = arcL^2,
     frce = U^2 * S,
     Mass_Spec_MetR = MetR / Mstarved,
     
     # convert to factor variables
     order = as.factor(as.character(order)),
     Treatment = as.factor(as.character(Treatment)),
     BeeID = as.factor(as.character(BeeID))
     )

# Create a new dataframe that calculates the changes for each individual bee
newDF <- data.frame()
colsTocalc = c("order", "M2", "MF", "MetR", "freq", "amp", "load", "MT", "perLoad", "frce", "arcL2", "freq2", "U", "arcL")
for(varb in colsTocalc){
     data_wide <- data.table::dcast(bdta, BeeID + Mstarved + S + Itspan + wingLen ~ 
                             Treatment, value.var=c(varb))
     colnames(data_wide)[6:7] = paste(varb, colnames(data_wide)[6:7], sep = "_")
     if(varb == colsTocalc[1]){
       newDF <- data_wide   
     }
     else newDF <- merge(newDF, data_wide, all.y = TRUE)    %>% as.tbl()
}

head(newDF)




# Calculate $\Delta$ variables
newDF <- newDF %>%
  mutate(
     deltaPercLoad = perLoad_H - perLoad_L,
     avgPercLoad = (perLoad_H + perLoad_L) / 2,
     deltaMetR = MetR_H - MetR_L,
     deltaFrq2 = freq2_H - freq2_L,
     deltaArcL = arcL_H - arcL_L,
     deltaArcL2 = arcL2_H - arcL2_L,
     deltaFreq2Perc = deltaFrq2 / deltaPercLoad,
     deltaLoad = scale(load_H - load_L, center = TRUE, scale = FALSE),
     dLoad_nonCent = load_H - load_L,
     deltaLoad2 = deltaLoad^2, 
     
     # reformat order so that it is more interpretable
     order_1 = recode(.$order_H, 
                      "2" = "loadedSecond", 
                      "1" = "loadedFirst")
    )


# Use PCA to combine 3 bee size measurements into a single predictor
aa = prcomp(newDF[, c("Mstarved", "Itspan", "S", "wingLen")], center = TRUE, scale = TRUE)
summary(aa) # 1st pc explains ~95% of the variance in the three measurements of size



# note, I changed the signs of the predictions so that higher PC1 values 
# correspond to bigger bees
p1 = -predict(aa)[,1] 

# add PC1 scores to dataset
newDF <- newDF %>%
  mutate(size_pc1 = p1, 
         size_pc1_2 = size_pc1^2)
# add PC1 to bdta
bdta <- bdta %>%
  full_join(newDF[,c("BeeID", "size_pc1", "size_pc1_2") ], 
            by = c("BeeID"))

```

---
# Visualize
---

```{r}

biplot(aa, main = "Three size measurements") # shows that all three size measurement are correlated

# show scatterplot matrix to see correlations among size predictors
car::scatterplotMatrix(newDF[, c("Mstarved", "Itspan", "S",  "wingLen",  "size_pc1")])

```





---
# Modeling & Figure generation
---

## Metabolic Rate vs. loading conditions
```{r}
m1 <- lmer(MetR ~ Treatment * MT + (1|BeeID), data = bdta)
summary(m1)

m2 <- update(m1, .~. - Treatment:MT)
anova(m2, m1, test = "LRT")

#######################################################
## Metabolic Rate vs. loading conditions
#######################################################

# Plot metabolic rate vs. body mass + load
fig3a <- bdta %>%
  mutate(Trt2 = recode(.$Treatment, "H" = "Heavy load", 
                       "L" = "Light load")) %>%
ggplot(aes(x = Mstarved + load, y = MetR )) + 
  geom_point(aes(shape = Trt2)) + 
  stat_smooth(aes(color = Trt2), method = "lm", se = F, size = 0.5) + 
  labs(y = expression("Metabolic rate (mL CO" [2] %.% ~ "hr"^{-1}~ ")" ),
       x = "Total mass (g)") + 
  lims(x = c(0.05,0.4), y = c(2, 18)) + 
  theme(legend.title = element_blank(), 
        legend.background = element_rect(colour = "black", size = 0.3), 
        legend.position = c(0.8, 0.16), 
        legend.text = element_text(size = 6), 
        axis.text = element_text(size = 6), 
        axis.title = element_text(size = 6), 
        legend.key.size = unit(0.2, "cm")) + 
  scale_color_grey() + 
  scale_shape_manual(values = c(16,1))

fig3a

ggsave(filename = file.path(figDir, "fig3A.png"), width = 6.5/2, height = 2, 
        plot = fig3a,
       units = "in", dpi = 500)


# calculate R^2 for heavy
mod1 <- lm(MetR_H ~ I(Mstarved + load_H), data = newDF)

# diagnostics
par(mfrow = c(2,3))
plot(mod1, which = 1:6)
dev.off()
summary(mod1)

# R^2 for light
mod2 <-  lm(MetR_L ~ I(Mstarved + load_L), data = newDF)
summary(mod2)
# diagnostics
par(mfrow = c(2,3))
plot(mod2, which = 1:6)
dev.off()
```


## Mass-specific metabolic rate versus percent loading

```{r}
#######################################################
### Mass-specific metabolic rate versus percent loading
#######################################################

# Plot mass-specific metabolic rate vs. percent loading
noFig <- bdta %>%
  mutate(Trt2 = recode(.$Treatment, "H" = "Heavy load", 
                       "L" = "Light load")) %>%
ggplot(aes(x = perLoad, y = Mass_Spec_MetR )) + 
  geom_point(aes(shape = Trt2)) + 
  stat_smooth(aes(color = Trt2), method = "lm", se = F, size = 0.5) + 
  labs(y = expression("Mass Specific Met. rate (mL CO" 
                      [2] %.% ~ "hr"^{-1} %.% ~ "g"^{-1}~ ")" ),
       x = "Percent load") + 
  #lims(x = c(0.05,0.4), y = c(2, 18)) + 
  theme(legend.title = element_blank(), 
        legend.background = element_rect(colour = "black", size = 0.3), 
        legend.position = c(0.2, 0.8), 
        legend.text = element_text(size = 6), 
        axis.text = element_text(size = 6), 
        axis.title = element_text(size = 6), 
        legend.key.size = unit(0.2, "cm")) + 
  scale_color_grey() + 
  scale_shape_manual(values = c(16,1))
noFig


ggsave(filename = file.path(figDir, "Mass_spec_MetRate.png"), width = 6.5/2, height = 2, 
        plot = noFig,
       units = "in", dpi = 500)


# calculate R^2 for heavy
mod1 <- lm(Mass_Spec_MetR ~ perLoad, data = bdta[bdta$Treatment == "H", ])
# diagnostics
par(mfrow = c(2,3))
plot(mod1, which = 1:6)
dev.off()
summary(mod1)
with(bdta[bdta$Treatment == "H", ], {
  plot(Mass_Spec_MetR ~ perLoad, pch = 20)
  lines(x = perLoad, y = predict(mod1))
})


mod2 <- lm(Mass_Spec_MetR ~ perLoad, data =  bdta[bdta$Treatment == "L", ])
summary(mod2)
# diagnostics
par(mfrow = c(2,3))
plot(mod2, which = 1:6)
dev.off()

with(bdta[bdta$Treatment == "L", ], {
  plot(Mass_Spec_MetR ~ perLoad)
  lines(x = perLoad, y = predict(mod2))
})


```

## Force vs. body mass

```{r}

m1 <- lmer(frce ~ Treatment * MT + (1|BeeID), data = bdta)
summary(m1)

m2 <- update(m1, .~. - Treatment:MT)
anova(m2, m1, test = "LRT")


# Plot force vs. body mass + load
fig3b <- bdta %>%
  mutate(Trt2 = recode(.$Treatment, "H" = "Heavy load", 
                       "L" = "Light load")) %>%
ggplot(aes(x = Mstarved + load, y = frce )) + 
  geom_point(aes(shape = Trt2)) + 
  stat_smooth(aes(color = Trt2), method = "lm", se = F, size = 0.5) + 
  labs(y = expression("Translational force proxy (m"^4 %.% ~ "s"^{-2}~ ")" ),
       x = "Total mass (g)") + 
  lims(x = c(0.05,0.4)) + 
  theme(legend.title = element_blank(), 
        legend.background = element_rect(colour = "black", size = 0.3), 
        legend.position = c(0.8, 0.16), 
        legend.text = element_text(size = 6), 
        axis.text = element_text(size = 6), 
        axis.title = element_text(size = 6), 
        legend.key.size = unit(0.2, "cm")) + 
  scale_color_grey() + 
  scale_shape_manual(values = c(16,1)) + 
  scale_y_continuous(labels = scales::scientific, limits = c(0, 5e-3))

fig3b

ggsave(filename = file.path(figDir, "fig3B.png"), width = 6.5/2, height = 2, 
        plot = fig3b,
       units = "in", dpi = 500)


# calculate R^2 for heavy
mod1 <- lm(frce ~ MT, data = bdta[bdta$Treatment == "H", ])
# diagnostics
par(mfrow = c(2,3))
plot(mod1, which = 1:6)
dev.off()
(summ <- summary(mod1))
summ$r.squared

mod2 <- lm(frce ~ MT, data =  bdta[bdta$Treatment == "L", ])
summary(mod2)
summary(mod2)$r.squared
# diagnostics
par(mfrow = c(2,3))
plot(mod2, which = 1:6)
dev.off()

```



## Stroke Amplitude
```{r}
# make a full model with all two-way interactions
m1 <- lmer(amp ~  (size_pc1 +  order + Treatment)^2 + (1|BeeID), data = bdta)

summary(m1)


### LRT's for interactions
m2.0 <- update(m1, .~. - order:Treatment )

anova(m1, m2.0)
summary(m2.0)


m2.1 <- update(m2.0,.~. - size_pc1:order)
anova(m2.0, m2.1)
summary(m2.1)


m2.2 <- update(m2.1, .~. - size_pc1:Treatment)
anova(m2.1, m2.2) ## drop all interactions

summary(m2.2)


# renaming model to simplify later typing
m2 <- m2.2

##### LRTs for main effects
## size
m3 <- update(m2, .~. - size_pc1)
anova(m2, m3, test = "Chi") # drop size_pc1
summary(m3)

# trt order
m4 <- update(m3, .~. - order)
anova(m3, m4) # keep order

# LRT for Treatment (load)
m5 <- update(m3, .~. - Treatment)
anova(m3, m5) # keep treatment

# summarize final model for paper
summary(m3)

# write output
summary(m3)$coefficients  
write.csv(summary(m3)$coefficients, file.path(dataOut, "AmpCoefs_TrtULVsL.csv" ))


# diagnostics
# qq plot
qqnorm(resid(m3), main = "")
qqline(resid(m3)) # ok

# residual plot
plot(fitted(m3), resid(m3), xlab = "fitted", ylab = "residuals")
abline(0,0)


# QQPlot for group-level effects
qqnorm(ranef(m3)$BeeID[[1]], main="Normal Q-Q plot for random effects")
qqline(ranef(m3)$BeeID[[1]]) # looks good

infl <- influence(m3, obs = TRUE)
plot(infl, which = 'cook')

bdta %>%
  mutate(Trt2 = recode(.$Treatment, "H" = "Heavy load", 
                       "L" = "Light load"), 
         order2 = recode(order, "1" = "First trial", 
                                  "2" = "Second trial")) %>%
ggplot(aes(x = order2, y = amp)) + 
  geom_point() + 
  facet_grid(~Trt2) + 
  geom_point(aes(x = order2, y = predict(m3, re.form = NA)), 
             color = "red", shape = 18, 
             size = 5) + 
  labs(x = "Trial num", y="Stroke amplitude")

```




## Wingbeat Freq

```{r}
# make a full model with all two-way interactions
m1 <- lmer(freq ~  (size_pc1 +  order + Treatment)^2 + (1|BeeID), data = bdta)

summary(m1)


### LRT's for interactions
m2.0 <- update(m1, .~. - size_pc1:Treatment )

anova(m1, m2.0)
summary(m2.0)


m2.1 <- update(m2.0,.~. - size_pc1:order)
anova(m2.0, m2.1)
summary(m2.1)


m2.2 <- update(m2.1, .~. - order:Treatment)
anova(m2.1, m2.2) ## drop all interactions

summary(m2.2)


# renaming model to simplify later typing
m2 <- m2.2

##### LRTs for main effects
## Treatment Order
m3 <- update(m2, .~. - order)
anova(m2, m3, test = "Chi")

# LRT for size
m4 <- update(m2, .~. - size_pc1)
anova(m2, m4)

# LRT for Treatment (load)
m5 <- update(m2, .~. - Treatment)
anova(m2, m5)

# summarize final model for paper
summary(m2)

# write output
summary(m2)$coefficients  
write.csv(summary(m2)$coefficients, file.path(dataOut, "FreqCoefs_TrtULVsL.csv" ))



# diagnostics
# qq plot
qqnorm(resid(m2), main = "")
qqline(resid(m2)) # good

# residual plot
plot(fitted(m2), resid(m2), xlab = "fitted", ylab = "residuals")
abline(0,0)


# QQPlot for group-level effects
qqnorm(ranef(m2)$BeeID[[1]], main="Normal Q-Q plot for random effects")
qqline(ranef(m2)$BeeID[[1]]) # looks good


infl <- influence(m2, obs = TRUE)
plot(infl, which = 'cook')

bdta %>%
  mutate(Trt2 = recode(.$Treatment, "H" = "Heavy load", 
                       "L" = "Light load"), 
         order2 = recode(order, "1" = "First trial", 
                                  "2" = "Second trial")) %>%
ggplot(aes(x = order2, y = freq)) + 
  geom_point() + 
  facet_grid(~Trt2) + 
  labs(x = "Trial num", y="Wingbeat Frequency")

```



# Investigate how Δ load, bee size, and treatment order affected changes in arc length2 between light and heavy trials

```{r}
# fit full model
m1 <- lm(deltaArcL2 ~  (deltaLoad +  size_pc1 + order_1)^2  + 
              size_pc1_2  + deltaLoad2, data = newDF)
summary(m1)

m2 <- update(m1, .~. - size_pc1:deltaLoad)
anova(m1, m2)

summary(m2)
m3 <- update(m2, .~. - deltaLoad:order_1)

anova(m2, m3)
summary(m3)

m4 <- update(m3, .~. - size_pc1:order_1)
anova(m3, m4) 
summary(m4)

m5 <- update(m4, .~. - deltaLoad2)
anova(m4,m5) 

summary(m5)

m6 <- update(m5, .~.  - size_pc1_2)
anova(m6, m5) 
summary(m6)

m7 <- update(m6, .~. - order_1)
anova(m7, m6)
summary(m7) # final model for paper

# Note that the t-stat^2 is the same as a partial f-test
anova(update(m7, .~. - deltaLoad), m7, test = "F") # p-value for delta load
anova(update(m7, .~. - size_pc1),  m7) # p-value for size

summary(m7)

### model visualization
# calculate partial residuals for deltaLoad
# these are the  residuals, minus the effect of detlaLoad
y <- residuals(m7, type = 'partial')[, "deltaLoad"]

# plot partial residuals with base R plotting
plot(x = newDF$deltaLoad, y = y)

# double check to make sure the slope for partial residual plots are the 
# same as in the original regression
summary(lm(y ~ newDF$deltaLoad)) 

# this is what the raw data look like
plot(x = newDF$deltaLoad, y = newDF$deltaArcL2)

# y axis isn't easily interpretable
ggplot(newDF, aes(x= deltaLoad, y = y)) + 
     geom_point(shape = 17) + 
     labs(x = "delta load", y = "partial residuals for delta load \n i.e. delta load effect on delta arclen^2 \n while subtracing affect of bee size and order") + 
     stat_smooth(method = 'lm', se = FALSE, color = "black", lty = 2) 


fig4a <- ggplot(newDF, aes(x= dLoad_nonCent, y = y)) + 
  geom_point(shape = 17) + 
  xlim(c(0.03, 0.12)) + 
  labs(x = "delta load (g)", y = "Δ arcLeng^2 partial residual" ) + 
  stat_smooth(method = 'lm', se = FALSE, color = "black", lty = 2, size = 0.5) + 
  scale_y_continuous(labels = scales::scientific, limits = c(-1.5e-4, 1.5e-4)) + 
  theme(legend.title = element_blank(), 
        legend.background = element_rect(colour = "black", size = 0.3), 
        legend.position = c(0.8, 0.16), 
        legend.text = element_text(size = 6), 
        axis.text = element_text(size = 6), 
        axis.title = element_text(size = 6), 
        legend.key.size = unit(0.2, "cm"))  
fig4a

ggsave(filename = file.path(figDir, "fig4a.png"), width = 6.5/2, height = 2, 
        plot = fig4a,
       units = "in", dpi = 500)


```


# Investigate how Δ load, bee size, and treatment order affected changes in freq^2 between light and heavy trials 

```{r}
# fit full model
m1 <- lm(deltaFrq2 ~  (deltaLoad +  size_pc1 + order_1)^2  + 
              size_pc1_2  + deltaLoad2, data = newDF)
summary(m1)
car::vif(m1) # some serious multicollinearity

m2 <- update(m1, .~. - size_pc1:order_1)
anova(m1, m2)

summary(m2)
m3 <- update(m2, .~. - deltaLoad:size_pc1)

anova(m2, m3)
summary(m3)

m4 <- update(m3, .~. - deltaLoad:order_1)
anova(m3, m4) 
summary(m4)

m5 <- update(m4, .~. - size_pc1_2)
anova(m4,m5) 

summary(m5)

m6 <- update(m5, .~.  - deltaLoad2)
anova(m6, m5) 
summary(m6) # final model for paper

m7 <- update(m6, .~. - size_pc1)
anova(m7, m6) # p-values for size
m8 <- update(m6, .~. - deltaLoad)
anova(m6, m8) # p-value for deltaLoad
m9 <- update(m6, .~. - order_1)
anova(m6, m9) # p-value for order


### model diagnostics
par(mfrow = c(2,3))
plot(m6, which = 1:6) # no glaring violations
par(mfrow = c(1,1))

car::vif(m6) # vif is a little high

### model visualization
# calculate partial residuals for deltaLoad
# these are the  residuals, minus the effect of detlaLoad
y <- residuals(m6, type = 'partial')[, "deltaLoad"]

# plot partial residuals with base R plotting
plot(x = newDF$deltaLoad, y = y)

# double check to make sure the slope for partial residual plots are the 
# same as in the original regression
summary(lm(y ~ newDF$deltaLoad)) 

# this is what the raw data look like
plot(x = newDF$deltaLoad, y = newDF$deltaFrq2)


# plot with ggplot2
# plot raw data w/ ggplot
ggplot(newDF, aes(x= size_pc1, y = deltaFrq2, color = deltaLoad, shape = order_1)) + 
     geom_point()

ggplot(newDF, aes(x= deltaLoad, y = deltaFrq2, color = size_pc1, shape = order_1)) + 
     geom_point()


# y axis isn't easily interpretable


ggplot(newDF, aes(x= dLoad_nonCent, y = y)) + 
     geom_point(shape = 17) + 
  xlim(c(0.03, 0.12)) + 
     labs(x = "delta load", y = "partial residuals for delta load \n i.e. delta load effect on freq^2 \n while subtracing affect of bee size and order") + 
     stat_smooth(method = 'lm', se = FALSE, color = "black", lty = 2) + 
     scale_y_continuous(labels = scales::scientific, limits = c(-8e3, 8e3))


fig4b <- ggplot(newDF, aes(x= dLoad_nonCent, y = y)) + 
  geom_point(shape = 17) + 
  xlim(c(0.03, 0.12)) + 
  labs(x = "delta load (g)", y = "Δ Frequency^2 partial residual" ) + 
  stat_smooth(method = 'lm', se = FALSE, color = "black", lty = 2, size = 0.5) + 
  scale_y_continuous(labels = scales::scientific, limits = c(-8e3, 8e3)) + 
  theme(legend.title = element_blank(), 
        legend.background = element_rect(colour = "black", size = 0.3), 
        legend.position = c(0.8, 0.16), 
        legend.text = element_text(size = 6), 
        axis.text = element_text(size = 6), 
        axis.title = element_text(size = 6), 
        legend.key.size = unit(0.2, "cm"))  
fig4b

ggsave(filename = file.path(figDir, "fig4b.png"), width = 6.5/2, height = 2, 
        plot = fig4b,
       units = "in", dpi = 500)

# partial residuals for order

y <- residuals(m6, type = 'partial')[, "order_1"]

ggplot(newDF, aes(x= order_1, y = y)) + 
     geom_boxplot() + 
     labs(x = "order", 
          y = "partial residuals for order \n i.e. order effect on freq^2 \n while subtracing affect of bee size and load") + 
     stat_smooth(method = 'lm', se = FALSE)

# holding bee size and delta load constant, if a bee was loaded second, (confusing, huh?) then it would have a much lower freq^2 than if it was loaded first

summary(m6)
```

<hr>
### Takeaways from (delta freq^2) model:
1. Order, size, and deltaLoad are associated with a change in frequency^2.  We find no evidence of non-linear relationships or interactions.
2. Holding size and order constant, we find that a larger deltaLoad causes a decrease in deltaFrequency^2
3. Holding other variables constant, an increase in bee size is associated with a larger deltaFrequency^2.
4. Holding other variables constant, if the bee was loaded second, then they had a lower deltaFrequency^2 than if they were loaded first. 

<hr>


# we investigated the change in metabolic rate between light and heavy loading trials

```{r}
mm1 <- lm(deltaMetR ~ deltaFrq2 + deltaArcL2 + deltaLoad + size_pc1 + deltaLoad2, data = newDF)
car::vif(mm1)

summary(mm1)

mm2 <- update(mm1, .~. - size_pc1)
anova(mm1, mm2)
summary(mm2)

mm3 <- update(mm2, .~. - deltaArcL2)
anova(mm3, mm2) # p-value for acrlength
summary(mm3)

mm4 <- update(mm3, .~. - deltaLoad2)
anova(mm3, mm4)
summary(mm4)

mm5 <- update(mm4, .~. - deltaLoad) 
anova(mm5, mm4) # p-value for load

mm6 <- update(mm4, .~. - deltaFrq2)
anova(mm4, mm6) # p-value for deltafrq2

# rename model
mm7 <- mm4

summary(mm7) # final model for paper


# calculate partial residuals from model 7 for deltaload and deltafreq2

prDload <- residuals(mm7, type = 'partial')[, "deltaLoad"]
prDfreq2 <- residuals(mm7, type = 'partial')[, "deltaFrq2"]


# plot partial residuals with base R plotting
plot(x = newDF$deltaLoad, y = prDload)
plot(x = newDF$deltaFrq2, y = prDfreq2)

partResidFig4 <- data.frame(deltaLoad = newDF$deltaLoad, 
                            partResDeltaLoad = prDload, 
                            deltaFreq2 = newDF$deltaFrq2, 
                            partResDeltaFreq2 = prDfreq2)


## model diagnostics
car::vif(mm7)

par(mfrow = c(2,3))
plot(mm7, which = 1:6) # no glaring violations
par(mfrow = c(1,1))



### model visualization
# calculate partial residuals for deltaLoad
# these are the  residuals, minus the effect of detlaLoad
y <- residuals(mm7, type = 'partial')[, "deltaLoad"]

# plot partial residuals with base R plotting
plot(x = newDF$deltaLoad, y = y)

# double check to make sure the slope for partial residual plots are the 
# same as in the original regression
summary(lm(y ~ newDF$deltaLoad)) 
summary(mm7)

# this is what the raw data look like
plot(x = newDF$deltaLoad, y = newDF$deltaMetR)


fig5a <- ggplot(newDF, aes(x= dLoad_nonCent, y = y)) + 
  geom_point(shape = 17) + 
  xlim(c(0.03, 0.12)) + 
  labs(x = "delta load (g)", y = "Δ Metabolic Rate partial residual" ) + 
  stat_smooth(method = 'lm', se = FALSE, color = "black", lty = 2, size = 0.5) + 
  scale_y_continuous(limits = c(-2, 3.5)) + 
  theme(legend.title = element_blank(), 
        legend.background = element_rect(colour = "black", size = 0.3), 
        legend.position = c(0.8, 0.16), 
        legend.text = element_text(size = 6), 
        axis.text = element_text(size = 6), 
        axis.title = element_text(size = 6), 
        legend.key.size = unit(0.2, "cm"))  
fig5a

ggsave(filename = file.path(figDir, "fig5a.png"), width = 6.5/2, height = 2, 
        plot = fig5a,
       units = "in", dpi = 500)




# these are the  residuals, minus the effect of delta freq^2
y <- residuals(mm7, type = 'partial')[, "deltaFrq2"]

# plot partial residuals with base R plotting
plot(x = newDF$deltaFrq2, y = y)

# double check to make sure the slope for partial residual plots are the 
# same as in the original regression
summary(lm(y ~ newDF$deltaFrq2)) 
summary(mm7)

# this is what the raw data look like
plot(x = newDF$deltaFrq2, y = newDF$deltaMetR)


fig5b <- ggplot(newDF, aes(x= deltaFrq2, y = y)) + 
  geom_point(shape = 17) + 
  xlim(c(-6*10^3, 9e3)) + 
  labs(x = "delta frequency^2", y = "Δ Metabolic Rate partial residual" ) + 
  stat_smooth(method = 'lm', se = FALSE, color = "black", lty = 2, size = 0.5) + 
  scale_y_continuous(limits = c(-3, 5)) + 
  theme(legend.title = element_blank(), 
        legend.background = element_rect(colour = "black", size = 0.3), 
        legend.position = c(0.8, 0.16), 
        legend.text = element_text(size = 6), 
        axis.text = element_text(size = 6), 
        axis.title = element_text(size = 6), 
        legend.key.size = unit(0.2, "cm"))  
fig5b

ggsave(filename = file.path(figDir, "fig5b.png"), width = 6.5/2, height = 2, 
        plot = fig5b,
       units = "in", dpi = 500)


```

<hr>
### Takeaways for deltaMetRate model:
1. Deltafreq^2 and deltaLoad are both associated with deltaMetR.
2. Holding deltafreq constant (not necessarily at 0), an increase in deltaload is associated with an increase in deltametabolicRate
3. Holding deltaLoad constant (not holding load constant, and again, not holding deltaload necessarily at 0), an increase in deltaFreq^2 is associated with an increase in deltaMetabolic rate. 
4. We found no evidence to suggest that a change in deltaArcL^2 is associated with deltametabolicRate. (this is not saying that arcLength doesn't affect metabolic rate).

<hr>

#  modeled the change in metabolic rate per 1% of additional loading

```{r}
newDF <- within(newDF, {dmr_dpl = deltaMetR / deltaPercLoad})
plot(deltaMetR  ~ dmr_dpl, data = newDF)

mod1 <- lm(dmr_dpl  ~ avgPercLoad + order_1 + size_pc1, data = newDF)
summary(mod1)

mod2 <- update(mod1, .~. - size_pc1)
anova(mod1, mod2) # remove size_pc1

mod3 <- update(mod2, .~. - order_1)
anova(mod2, mod3) # remove order


summary(mod3) # final model for paper

# get p-values for avgPercLoad and order
anova(mod2, update(mod3, .~. - avgPercLoad)) # p-value for avg perc load

# plot(I(deltaMetR)  ~ avgPercLoad, data = newDF)
# plot(I(deltaMetR / avgPercLoad)  ~ deltaMetR, data = newDF)

car::scatterplotMatrix(newDF[, c("avgPercLoad", "deltaMetR", "size_pc1")])
car::vif(mod1)  # looks fine


par(mfrow = c(2,2))
plot(mod3) # possible nonlinear trend in residuals

plot(mod3, which = 4)

# update model to add a non-linear term
newDF <- within(newDF, {avgPercLoad_cent = as.numeric(scale(newDF$avgPercLoad, center = TRUE, scale = FALSE))})
newDF$apl_cent2 <- with(newDF, avgPercLoad_cent^2)

m11 <- lm(dmr_dpl  ~ avgPercLoad_cent  + apl_cent2 + order_1 + size_pc1, data = newDF)
summary(m11)
car::vif(m11) # vif is much better with centered variables

m11a <- update(m11, .~. - size_pc1)
anova(m11, m11a) # remove size_pc1


m11b <- update(m11a, .~. - apl_cent2)
anova(m11a, m11b) # drop squared term
summary(m11b)

m11c <- update(m11b, .~. -order_1 )
anova(m11b, m11c) # remove order_1

summary(m11c) # final model for paper

## visualize model for deltametRate/avgPercLoading

par(mfrow = c(1,1))
plot(dmr_dpl  ~ avgPercLoad, col = factor(order_1), data = newDF, pch = 20)

ndf <- data.frame(avgPercLoad_cent = seq(min(newDF$avgPercLoad_cent), max(newDF$avgPercLoad_cent), length.out = 200), 
                  order_1 = as.factor(rep(c("loadedFirst", "loadedSecond"), 100)))

ndf$apl_cent2 <- ndf$avgPercLoad_cent^2

ndf$avgPercLoad <- ndf$avgPercLoad_cent + mean(newDF$avgPercLoad)
predDF <- data.frame(preds = predict(mod3, newdata = ndf), ndf)


newDF$order_1

newDF %>%
  mutate(load_order = recode(order_1, "loadedFirst" = "Heavy -> Light",
                             "loadedSecond" = "Light -> Heavy")) %>%
ggplot( aes(x = avgPercLoad, y = dmr_dpl)) + 
     geom_point(aes(color = load_order, shape = load_order)) + 
     geom_line(data = predDF, aes(x = avgPercLoad, y = preds)) + 
     labs(x = "Average Load (% bodymass)", 
          y = "Change in Metabolic Rate (mL CO2 / hr) / \n Change in Load (% bodymass)") + 
     scale_color_viridis_d( name = "Order", end = 0.8) + 
  theme(legend.title = element_blank(), 
        legend.background = element_rect(colour = "black", size = 0.3), 
        legend.position = c(0.8,0.8), 
        legend.text = element_text(size = 6), 
        axis.text = element_text(size = 6), 
        axis.title = element_text(size = 6), 
        legend.key.size = unit(0.2, "cm"))  + 
  scale_shape_manual(name = "Order", values = c(25, 24))

ggsave(filename = file.path(figDir, "fig6a.png"), width = 6.5/2, height = 2, 
       units = "in", dpi = 500)

```


# (DeltaFrq^2/deltaPercLoad) ~ avgPercLoad + order_1 + size_pc1

```{r}
newDF <- within(newDF, {df2_dpl = deltaFrq2 / deltaPercLoad})

mod1_f <- lm(deltaFrq2  ~  avgPercLoad + order_1 + size_pc1, data = newDF)

summary(mod1_f)

mod2_f <- update(mod1_f, .~. - size_pc1)
anova(mod1_f, mod2_f) # remove size_pc1

summary(mod2_f) # not final model for paper

# get p-values for avgPercLoad 
anova(mod2_f,  update(mod2_f, .~. - order_1)) # p-value for order

par(mfrow = c(2,2))
plot(mod2_f) # non-linearity in residuals

plot(mod2_f, which = 4)


# update model to add a non-linear term
m22 <- lm(df2_dpl  ~ avgPercLoad_cent  + apl_cent2 + order_1 + size_pc1, data = newDF)
summary(m22)
car::vif(m22) # vif is much better with centered variables

m22a <- update(m22, .~. - size_pc1)
anova(m22, m22a) # remove size_pc1

summary(m22a)
par(mfrow = c(2,2))
plot(m22a) # residuals look better, though slight fan shape

plot(m22a, which = 4) ## row 22 looks highly influential
nd22 <- newDF[-22, ]
m22s <- lm(df2_dpl  ~ avgPercLoad_cent  + apl_cent2 + order_1, data = nd22)
summary(m22s) # no major change when we remove obs num 22
plot(m22s, which = 4)

summary(m22a) # final model for paper 

## visualize model for deltametRate/avgPercLoading

ndf <- data.frame(avgPercLoad_cent = seq(min(newDF$avgPercLoad_cent), max(newDF$avgPercLoad_cent), length.out = 200), 
                  order_1 = as.factor(rep(c("loadedFirst", "loadedSecond"), 100)))

ndf$apl_cent2 <- ndf$avgPercLoad_cent^2

ndf$apl <- ndf$avgPercLoad_cent + mean(newDF$avgPercLoad)


predDF <- data.frame(preds = predict(m22a, newdata = ndf, se = TRUE), ndf)
predDF<- mutate(predDF, 
                            load_order = recode(order_1, 
                                                "loadedFirst" = "Heavy -> Light",
                             "loadedSecond" = "Light -> Heavy"))

par(mfrow = c(1,1))
plot(df2_dpl  ~ avgPercLoad, col = factor(order_1), data = newDF, pch = 20)

newDF[22, ]

ggplot(newDF, aes(x = avgPercLoad, y = df2_dpl)) + 
     geom_point(aes(color = order_1),shape = 17) + 
     geom_point(aes(size = BeeID == "E42")) +  # show the influential point
     geom_line(data = predDF, aes(x = apl, y = preds.fit, color = order_1)) + 
     # geom_line(data = predDF, aes(x = apl, y = preds.fit + 1.96*preds.se.fit, color = order_1), lty = 2) + 
     # geom_line(data = predDF, aes(x = apl, y = preds.fit - 1.96*preds.se.fit, color = order_1), lty = 2) + 
     labs(x = "Average Load (% bodymass)", 
          y = "Change in wingbeat freq^2 (hz^2) / \n Change in Load (% bodymass)") + 
     scale_color_viridis_d( name = "Order",end = 0.8) + 
     theme(legend.position = c(0.8,0.8)) 


newDF %>%
  mutate(load_order = recode(order_1, "loadedFirst" = "Heavy -> Light",
                             "loadedSecond" = "Light -> Heavy")) %>%
ggplot( aes(x = avgPercLoad, y = df2_dpl)) + 
     geom_point(aes(color = load_order, shape = load_order )) + 
     geom_line(data = predDF, aes(x = apl, y = preds.fit, color = load_order, lty = load_order)) + 
     labs(x = "Average Load (% bodymass)", 
          y = "Change in wingbeat freq^2 (hz^2) / \n Change in Load (% bodymass)") + 
     scale_color_viridis_d( name = "Order", end = 0.8) + 
     theme(legend.title = element_blank(), 
        legend.background = element_rect(colour = "black", size = 0.3), 
        legend.position = c(0.8,0.8), 
        legend.text = element_text(size = 6), 
        axis.text = element_text(size = 6), 
        axis.title = element_text(size = 6), 
        legend.key.size = unit(0.2, "cm"))  + 
  scale_shape_manual(name = "Order", values = c(25, 24)) +
  scale_linetype_manual(name = "Order", values = c(2,1))

ggsave(filename = file.path(figDir, "fig6b.png"), width = 6.5/2, height = 2, 
       units = "in", dpi = 500)
```



# (DeltaArcL^2/deltaPerLoad) ~ avgPercLoad + order_1 + size_pc1

```{r}
newDF <- within(newDF, {da2_dpl = deltaArcL2 / deltaPercLoad})

mod1_a <- lm(da2_dpl  ~ avgPercLoad + order_1 + size_pc1, data = newDF)
summary(mod1_a) # P-values for original model

mod2_a <- update(mod1_a, .~. - avgPercLoad)  # I also checked for a squared term in this model (code not shown)
anova(mod1_a, mod2_a) # remove avgPercLoad

summary(mod2_a)


mod2_b <- update(mod2_a, .~. - size_pc1)
anova(mod2_a, mod2_b)  # remove size

mod2c <- update(mod2_b, .~. - order_1)
anova(mod2c, mod2_b) # remove order

summary(mod2c) # final mod for paper

par(mfrow = c(2,2))
plot(mod2_a)

plot(mod2_a, which = 4)

# visualize model
ndf <- data.frame(avgPercLoad_cent = seq(min(newDF$avgPercLoad_cent), max(newDF$avgPercLoad_cent), length.out = 200), 
                  order_1 = as.factor(rep(c("loadedFirst", "loadedSecond"), 100)))

ndf$apl_cent2 <- ndf$avgPercLoad_cent^2

ndf$avgPercLoad <- ndf$avgPercLoad_cent + mean(newDF$avgPercLoad)

predDF <- data.frame(preds = predict(mod2c, newdata = ndf, se = TRUE), ndf)

par(mfrow = c(1,1))
plot(da2_dpl  ~ avgPercLoad, col = factor(order_1), data = newDF, pch = 20)

newDF %>%
  mutate(load_order = recode(order_1, "loadedFirst" = "Heavy -> Light",
                             "loadedSecond" = "Light -> Heavy")) %>%
ggplot(aes(x = avgPercLoad, y = da2_dpl)) + 
     geom_point(aes(color = load_order, shape = load_order)) + 
     labs(x = "Average Load (% bodymass)", 
          y = "Change in arc length^2 (radians^2) / \n Change in Load (% bodymass)") +     
     ylim(c(0, 5e-6)) + 
     scale_color_viridis_d( name = "Order", end = 0.8) + 
     theme(legend.title = element_blank(), 
        legend.background = element_rect(colour = "black", size = 0.3), 
        legend.position = c(0.8,0.8), 
        legend.text = element_text(size = 6), 
        axis.text = element_text(size = 6), 
        axis.title = element_text(size = 6), 
        legend.key.size = unit(0.3, "cm"))  + 
    scale_shape_manual(name = "Order", values = c(25, 24)) 

ggsave(filename = file.path(figDir, "fig6c.png"), width = 6.5/2, height = 2, 
       units = "in", dpi = 500)


```



# Session Info
```{r}
sessionInfo()

Sys.time()

```